### 2019-12-11 README.Rmd
Initial readme changes

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

# msBalance

## Overview

msBalance is a an R package for analysis of balance responses in people with and without multiple sclerosis (MS)

## Downloading from GitHub for development

To work on the package for others to use:

1. Be sure to have `devtools` installed.
1. Use RStudio to create a local copy in `~/tmp` based on https://github.com/jlucasmckay/MSbalance
1. `devtools::check()`
1. `devtools::build_vignettes()`
1. `browseVignettes("MSbalance")`
1. Edit
1. Stage changed files, commit, and push changes with the git panel.

## Installing from GitHub for use

Currently msBalance is available on github.
It is being developed to meet CRAN requirements but is not available there at present.
To use the package yourself:

1. Be sure to have `devtools` installed.
1. Call `remove.packages("msBalance")` if you have installed the package previously.
1. `devtools::install_github("jlucasmckay/msBalance", build_vignettes = TRUE)`
1. `browseVignettes("msBalance")`

### 2019-12-11 README.Rmd
updated README

@@
 
-msBalance is a an R package for analysis of balance responses in people with and without multiple sclerosis (MS)
+msBalance is a an R package for analysis of balance responses in people with and without multiple sclerosis (MS).
+The source data were created by Jessie Huisinga, Ph.D., and collected by Daniel Peterson, Ph.D.
+The analysis code was primarily developed by J. LUcas McKay, Ph.D., M.S.C.R.
 
-## Downloading from GitHub for development
+## Getting msBalance
+
+### Downloading from GitHub for development
 
@@
 
-## Installing from GitHub for use
+### Installing from GitHub for use
 
@@
 1. `browseVignettes("msBalance")`
+
+## Using msBalance
+
+The best way to get started is by using the `msBalance` vignette.

### 2019-12-12 R/dataImport.R
first function prototypes

@@ -1,0 +1,4 @@
+#' download source data files from remote location
+#'
+#' @export
+downloadRemoteData = function() NULL

### 2019-12-12 README.Rmd
updated vignettes and documentation
updated docs

@@
 The source data were created by Jessie Huisinga, Ph.D., and collected by Daniel Peterson, Ph.D.
-The analysis code was primarily developed by J. LUcas McKay, Ph.D., M.S.C.R.
+The analysis code was primarily developed by J. Lucas McKay, Ph.D., M.S.C.R.
 
-## Getting msBalance
+## Getting and installing msBalance
 
@@
 
-To work on the package for others to use:
+To download, develop, and install:
 
@@
 1. Use RStudio to create a local copy in `~/tmp` based on https://github.com/jlucasmckay/MSbalance
-1. `devtools::check()`
-1. `devtools::build_vignettes()`
-1. `browseVignettes("MSbalance")`
-1. Edit
+1. Edit vignettes/msBalance.Rmd and R files in R/.
 1. Stage changed files, commit, and push changes with the git panel.
-
-### Installing from GitHub for use
-
-Currently msBalance is available on github.
-It is being developed to meet CRAN requirements but is not available there at present.
-To use the package yourself:
-
-1. Be sure to have `devtools` installed.
-1. Call `remove.packages("msBalance")` if you have installed the package previously.
 1. `devtools::install_github("jlucasmckay/msBalance", build_vignettes = TRUE)`
-1. `browseVignettes("msBalance")`
-
-## Using msBalance
-
-The best way to get started is by using the `msBalance` vignette.
+1. `vignette("msBalance")`

### 2019-12-12 README.Rmd
updated docs

@@
 
-To download, develop, and install:
-
 1. Be sure to have `devtools` installed.
-1. Use RStudio to create a local copy in `~/tmp` based on https://github.com/jlucasmckay/MSbalance
+1. Use RStudio to create a local copy in `~/tmp` based on https://github.com/jlucasmckay/msBalance
+    1. In RStudio, start a new Project: `File > New Project > Version Control > Git`
+    1. In the “repository URL” paste: https://github.com/jlucasmckay/msBalance
+    1. On OS X, make under `tmp`
+    1. Click “Create Project” to create a new directory, which will be:
+        * A directory/folder under `tmp`
+        * A Git repository, linked to the remote GitHub repository
+        * An RStudio Project
+    1. If there are files that need to be added to the remote, they can be placed in this folder or dragged / dropped into the GitHub interface.
 1. Edit vignettes/msBalance.Rmd and R files in R/.
 1. Stage changed files, commit, and push changes with the git panel.
+
+### Installing from GitHub for use
 1. `devtools::install_github("jlucasmckay/msBalance", build_vignettes = TRUE)`

### 2019-12-12 m/loadData.m
added supplemental matlab code

@@ -1,0 +1,25 @@
+% loadData.m
+% 2019 10 24
+% JLM
+
+% free up memory
+clear
+
+% load data, which is organized as a table T
+dataDir = 
+load("./data/AllConsolidatedData_CS.mat")
+c = T
+load("./data/AllConsolidatedData_MS.mat")
+m = T
+
+% load data dictionary
+vars = readtable("./data/dataDictionary.csv")
+idVars = vars.name(vars.class == "id" | vars.class == "dem")
+timeVars = vars.name(vars.class == "time")
+
+% pipe the patient-level-data to two csv files that can be combined in R
+writetable(m(:,ismember(m.Properties.VariableNames,idVars)), './data/m.csv')
+writetable(c(:,ismember(c.Properties.VariableNames,idVars)), './data/c.csv')
+
+% free up memory
+clear

### 2019-12-12 m/loadData.m
added matlab files

@@ -1,3 +1,4 @@
+% /Applications/MATLAB_R2019a.app/bin/matlab -nojvm -nodesktop -nosplash
 % loadData.m
 % 2019 10 24
 % JLM
@@ -5,21 +6,23 @@
 % free up memory
 clear
 
-% load data, which is organized as a table T
-dataDir = 
-load("./data/AllConsolidatedData_CS.mat")
+% specify data directory
+dataDirectory = "/Users/lucasmckay/OneDrive - Emory University/2019 Peterson MS EMG/data/"
+
+% load data for CS and MS subjects, which are organized as tables
+load(dataDirectory + "AllConsolidatedData_CS.mat")
 c = T
-load("./data/AllConsolidatedData_MS.mat")
+load(dataDirectory + "AllConsolidatedData_MS.mat")
 m = T
 
 % load data dictionary
-vars = readtable("./data/dataDictionary.csv")
+vars = readtable(dataDirectory + "dataDictionary.csv")
 idVars = vars.name(vars.class == "id" | vars.class == "dem")
 timeVars = vars.name(vars.class == "time")
 
 % pipe the patient-level-data to two csv files that can be combined in R
-writetable(m(:,ismember(m.Properties.VariableNames,idVars)), './data/m.csv')
-writetable(c(:,ismember(c.Properties.VariableNames,idVars)), './data/c.csv')
+writetable(m(:,ismember(m.Properties.VariableNames,idVars)), dataDirectory + "m.csv")
+writetable(c(:,ismember(c.Properties.VariableNames,idVars)), dataDirectory + "m.csv")
 
 % free up memory
 clear

### 2019-12-12 README.Rmd
new location for csv swap

@@
 
+### Navigating a mixture of Matlab and R code
+
+The source files are in .mat format, using structures like tables and other things that cannot be encoded in the earlier V4-6 .mat file types that can be easily read by R.
+There is a mixture of matlab preprocessing code (files in `m/`) into .csv files that R can then read (files in `R/`).
+A directory `csv/` is created to hold these files.
+The contents of this directory are ignored by git so that these files are not automatically loaded to GitHub, for security purposes.
+File path locations need to be
+
 ### Installing from GitHub for use
+
 1. `devtools::install_github("jlucasmckay/msBalance", build_vignettes = TRUE)`

### 2019-12-19 R/msBalance.Rmd
reorganized project

---
title: "msBalance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{msBalance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Locate the Matlab source files.
Locate the source data files on your computer.
These are intended to be kept in 
This will need to be modified.

```{r}
suppressMessages(library(R.matlab))

path = "~/OneDrive - Emory University/2019 Peterson MS EMG/data"
(filePathMS = file.path(path, "AllConsolidatedData_MS.mat"))
(filePathCS = file.path(path, "AllConsolidatedData_CS.mat"))

dataMS = read.mat(filePathMS)
dataCS = read.mat(filePathCS)
```

### 2019-12-19 README.Rmd
added paths.R

@@
 
-msBalance is a an R package for analysis of balance responses in people with and without multiple sclerosis (MS).
+msBalance is a an R project for analysis of balance responses in people with and without multiple sclerosis (MS).
 The source data were created by Jessie Huisinga, Ph.D., and collected by Daniel Peterson, Ph.D.
@@
     1. If there are files that need to be added to the remote, they can be placed in this folder or dragged / dropped into the GitHub interface.
-1. Edit vignettes/msBalance.Rmd and R files in R/.
+1. Edit msBalance.Rmd and R files in R/.
 1. Stage changed files, commit, and push changes with the git panel.
@@
 The contents of this directory are ignored by git so that these files are not automatically loaded to GitHub, for security purposes.
-File path locations need to be
-
-### Installing from GitHub for use
-
-1. `devtools::install_github("jlucasmckay/msBalance", build_vignettes = TRUE)`
-1. `vignette("msBalance")`
+File path locations need to be manually changed in R/paths.R

### 2019-12-19 R/dataImport.R
more changes to project from package

@@ -1,4 +1,2 @@
-#' download source data files from remote location
-#'
-#' @export
+# download source data files from remote location
 downloadRemoteData = function() NULL

### 2019-12-19 R/msBalance.Rmd
updated main file

@@
 ## Locate the Matlab source files.
-Locate the source data files on your computer.
-These are intended to be kept in 
-This will need to be modified.
 
 ```{r}
-suppressMessages(library(R.matlab))
-
-path = "~/OneDrive - Emory University/2019 Peterson MS EMG/data"
-(filePathMS = file.path(path, "AllConsolidatedData_MS.mat"))
-(filePathCS = file.path(path, "AllConsolidatedData_CS.mat"))
-
-dataMS = read.mat(filePathMS)
-dataCS = read.mat(filePathCS)
 ```

### 2019-12-19 R/dataImport.R

@@ -1,2 +1,4 @@
+
+
 # download source data files from remote location
 downloadRemoteData = function() NULL

### 2019-12-19 R/msBalance.Rmd
updated structures

@@
 ---
-title: "msBalance"
+title: "Balance responses in MS"
 output: rmarkdown::html_vignette
@@
 knitr::opts_chunk$set(
+  echo = FALSE,
   collapse = TRUE,
@@
 
-## Locate the Matlab source files.
+```{r}
+source("../paths.txt")
+```
 
-```{r}
-```
+## Locate sources
+Reading source files from `r dataDir`
+
+## Run Matlab code to load data

### 2019-12-19 README.Rmd
updated structures

--- 2019-12-12 README.Rmd
@@
     1. If there are files that need to be added to the remote, they can be placed in this folder or dragged / dropped into the GitHub interface.
-1. Edit vignettes/msBalance.Rmd and R files in R/.
+1. Path locations (used by R and Matlab code) are in paths.txt
+1. The main analysis is in R/msBalance.Rmd
 1. Stage changed files, commit, and push changes with the git panel.
 
-### Installing from GitHub for use
-1. `devtools::install_github("jlucasmckay/msBalance", build_vignettes = TRUE)`
-1. `vignette("msBalance")`
+### Navigating a mixture of Matlab and R code
+
+The source files are in .mat format.
+There is a mixture of matlab preprocessing code (files in m/) into .csv files that R can then read (files in R/).
+A directory csv/ is created to hold these files.
+The contents of this directory are ignored by git so that these files are not automatically loaded to GitHub.

### 2019-12-19 README.Rmd
changed file extensions

@@
 
-msBalance is a an R project for analysis of balance responses in people with and without multiple sclerosis (MS).
+msBalance is an R project for analysis of balance responses in people with and without multiple sclerosis (MS).
 The source data were created by Jessie Huisinga, Ph.D., and collected by Daniel Peterson, Ph.D.
@@
 
-1. Be sure to have `devtools` installed.
-1. Use RStudio to create a local copy in `~/tmp` based on https://github.com/jlucasmckay/msBalance
+1. Use RStudio to create a local copy of the project
     1. In RStudio, start a new Project: `File > New Project > Version Control > Git`
     1. In the “repository URL” paste: https://github.com/jlucasmckay/msBalance
-    1. On OS X, make under `tmp`
-    1. Click “Create Project” to create a new directory, which will be:
-        * A directory/folder under `tmp`
-        * A Git repository, linked to the remote GitHub repository
-        * An RStudio Project
+    1. On OS X, you can make this under `tmp` - this is a temporary file that can be deleted later
     1. If there are files that need to be added to the remote, they can be placed in this folder or dragged / dropped into the GitHub interface.
-1. Path locations (used by R and Matlab code) are in paths.txt
-1. The main analysis is in R/msBalance.Rmd
+1. Path locations (used by Matlab code) are in `msBalance/paths.m`
+1. The main analysis is in `msBalance/R/msBalance.Rmd`
 1. Stage changed files, commit, and push changes with the git panel.
@@
 
-The source files are in .mat format.
-There is a mixture of matlab preprocessing code (files in m/) into .csv files that R can then read (files in R/).
-A directory csv/ is created to hold these files.
-The contents of this directory are ignored by git so that these files are not automatically loaded to GitHub.
+1. The source files are in `.mat` format.
+1. There is a mixture of matlab preprocessing code (files in `m/`) into `.csv` files that R can then read (files in `R/`).
+1. A directory `csv/` is created to hold these files. The contents of this directory are not automatically loaded to GitHub.

### 2019-12-19 m/loadData.m

@@ -6,8 +6,8 @@
 % free up memory
 clear
 
-% specify data directory
-dataDirectory = "/Users/lucasmckay/OneDrive - Emory University/2019 Peterson MS EMG/data/"
+% specify data directory - 
+run("../paths.m")
 
 % load data for CS and MS subjects, which are organized as tables
 load(dataDirectory + "AllConsolidatedData_CS.mat")

### 2019-12-19 m/loadData.m
updated path locations

@@ -6,14 +6,17 @@
 % free up memory
 clear
 
-% specify data directory - 
-run("../paths.m")
+% specify data directory
+dataDirectory = "/Users/lucasmckay/OneDrive - Emory University/2019 Peterson MS EMG/data/"
+
+% specify csv file directory
+csvDirectory = "../csv/"
 
 % load data for CS and MS subjects, which are organized as tables
 load(dataDirectory + "AllConsolidatedData_CS.mat")
-c = T
+c = T;
 load(dataDirectory + "AllConsolidatedData_MS.mat")
-m = T
+m = T;
 
 % load data dictionary
 vars = readtable(dataDirectory + "dataDictionary.csv")
@@ -21,8 +24,8 @@ idVars = vars.name(vars.class == "id" | vars.class == "dem")
 timeVars = vars.name(vars.class == "time")
 
 % pipe the patient-level-data to two csv files that can be combined in R
-writetable(m(:,ismember(m.Properties.VariableNames,idVars)), dataDirectory + "m.csv")
-writetable(c(:,ismember(c.Properties.VariableNames,idVars)), dataDirectory + "m.csv")
+writetable(m(:,ismember(m.Properties.VariableNames,idVars)), csvDirectory + "m.csv")
+writetable(c(:,ismember(c.Properties.VariableNames,idVars)), csvDirectory + "m.csv")
 
 % free up memory
 clear

### 2019-12-19 paths.m
neep

@@ -1,1 +1,0 @@
-dataDirectory = "/Users/lucasmckay/OneDrive - Emory University/2019 Peterson MS EMG/data/"

### 2019-12-19 m/loadData.m
more changes

@@ -7,10 +7,10 @@
 clear
 
 % specify data directory
-dataDirectory = "/Users/lucasmckay/OneDrive - Emory University/2019 Peterson MS EMG/data/"
+run(".."+filesep+"paths.m")
 
 % specify csv file directory
-csvDirectory = "../csv/"
+csvDirectory = ".."+filesep+"csv"+filesep
 
 % load data for CS and MS subjects, which are organized as tables
 load(dataDirectory + "AllConsolidatedData_CS.mat")
@@ -25,7 +25,7 @@ timeVars = vars.name(vars.class == "time")
 
 % pipe the patient-level-data to two csv files that can be combined in R
 writetable(m(:,ismember(m.Properties.VariableNames,idVars)), csvDirectory + "m.csv")
-writetable(c(:,ismember(c.Properties.VariableNames,idVars)), csvDirectory + "m.csv")
+writetable(c(:,ismember(c.Properties.VariableNames,idVars)), csvDirectory + "c.csv")
 
 % free up memory
 clear

### 2019-12-19 paths.m

@@ -1,0 +1,1 @@
+dataDirectory = "/Users/lucasmckay/OneDrive - Emory University/2019 Peterson MS EMG/data/"

### 2019-12-20 R/msBalance.Rmd
changed file extensions
more changes
Added bibliographic information, initial frequency counts.

@@
 title: "Balance responses in MS"
-output: rmarkdown::html_vignette
-vignette: >
-  %\VignetteIndexEntry{msBalance}
-  %\VignetteEngine{knitr::rmarkdown}
-  %\VignetteEncoding{UTF-8}
+author:
+- J. Lucas McKay
+- Daniel S. Peterson
+date: "`r format(Sys.time(), '%d %B %Y')`"
+output: html_document
+bibliography: bibliography.bib
+csl: nature.csl
 ---
@@
 ```{r}
-source("../paths.txt")
+source("../paths.m")
 ```
 
-## Locate sources
-Reading source files from `r dataDir`
+# Introduction
+This document describes the analysis of standing balance responses in people with and without multiple sclerosis (MS) using wavelet-based functional ANOVA (wfANOVA[@mckay2013]).
+MS has profound effects on balance,[@vanliew2019] including altering protective stepping behaviors.[@peterson2016]
+The main objective of this study was to examine the impact of MS on the pattern of magnitude and timing of muscle activation during feet-in-place balance responses.
 
-## Run Matlab code to load data
+# Methods and Results
+
+## Data sources
+```{r, message=FALSE, warning=FALSE}
+library(vroom)
+library(dplyr)
+library(stringr)
+library(lubridate)
+`%contains%` = function(s, pattern) grepl(pattern, s)
+`%omits%` = function(s, pattern) !grepl(pattern, s)
+
+# load participant-level information
+m = vroom("../csv/m.csv") %>% mutate(Group = "MS") %>% mutate(Age = round(age)) %>% mutate(Sex = toupper(gender)) %>% rename(Participant = Subj)
+c = vroom("../csv/c.csv") %>% mutate(Group = "Control") %>% mutate(Age = round(age)) %>% mutate(Sex = toupper(gender)) %>% rename(Participant = Subj)
+
+# concatenate the participants.
+note that some of them appear to be coded as follow-ups.
+this is very useful, but may not be needed for the first report.
+they appear to be coded as such by the presence of "F1", "F2" in the participant name.
+convert this to a visit number (1, 2, 3, 4)
+participants = full_join( m %>% select(-trial, -age, -gender) %>% distinct(), c %>% select(-trial, -age, -gender) %>% distinct() ) %>%
+  mutate(`Visit Number` = ifelse(Participant %omits% "F\\d", 1, as.numeric(str_extract(str_extract(Participant, "F\\d"),"\\d"))+1)) %>% 
+  mutate(`Visit Date` = date(dmy_hms(obsDate))) %>% select(-obsDate) %>% 
+  mutate(`Visit ID` = Participant) %>% 
+  mutate(`Visit Code` = str_extract(Participant, "F\\d")) %>% 
+  mutate(Participant = str_replace(Participant, "F\\d", "")) %>% 
+  mutate(Group = factor(Group, levels = c("MS","Control")))
+
+# identify enrollment date for each participant and calculate amount of observation time
+participants = participants %>% left_join(participants %>% group_by(Participant) %>% summarize(`Enrollment Date` = min(`Visit Date`, na.rm = TRUE))) %>% 
+  mutate(`Time Since Enrollment (Days)` = ifelse(Group == "MS", `Visit Date` - `Enrollment Date`, 0))
+
+# calculate amount of follow-up
+followUp = participants %>% select(Group, Participant, `Time Since Enrollment (Days)`) %>% group_by(Group, Participant) %>% summarize(`Study Visits Available (n)` = n(), `Follow-Up Available (Days)` = max(`Time Since Enrollment (Days)`))
+
+```
+
+This study performed a secondary analysis on existing electromyographic recordings during balance responses in people with and without MS.
+Individual subjects were as follows:
+
+```{r, warning=FALSE, message=FALSE}
+library(janitor)
+library(kableExtra)
+median0 = function (x) median(x[x>0], na.rm = TRUE)
+min0 = function (x) min(x[x>0], na.rm = TRUE)
+max0 = function (x) max(x[x>0], na.rm = TRUE)
+
+k = function (x) kable_styling(x, full_width = F, position = "left")
+t = followUp %>% tabyl(Group) %>% adorn_totals("row") %>% adorn_pct_formatting(digits = 0) %>% rename(N = n, `Percent`=percent)
+followUpAvailable = followUp %>% filter(Group=="MS") %>% ungroup() %>% mutate(f = `Follow-Up Available (Days)`>0) %>% summarize(fAvail = sum(f==TRUE,na.rm = TRUE)/n()) %>% pull() %>% scales::percent()
+followUpMed = median0(followUp$`Follow-Up Available (Days)`)
+followUpMin = min0(followUp$`Follow-Up Available (Days)`)
+followUpMax = max0(followUp$`Follow-Up Available (Days)`)
+kable(t) %>% k()
+```
+
+Of these, follow-up visits were available for several of the participants with MS (≈`r followUpAvailable`).
+Among the participants with MS and follow-up available, the median duration was `r followUpMed` days, and the range was from `r followUpMin` to `r followUpMax` days.
+The number of visits was distributed as follows:
+
+```{r, warning=FALSE, message=FALSE}
+library(janitor)
+library(kableExtra)
+t = followUp %>% tabyl(Group, `Study Visits Available (n)`)
+kable(t) %>% k() %>% add_header_above(c(" " = 1, "Number of Study Visits Available" = 4))
+```
+
+## Exclusions
+Beginning with N=TK records, TK were excluded (N=TK, MS, N=TK, Control) due to incomplete demographic information.
+Characteristics of the remaining participants included in analyses are summarized in Table 1.
+
+## Demographic and Clinical Characteristics
+
+# Disclosures
+
+## Funding sources and conflict of interest
+This study was supported in part by National Institutes of Health grant K25HD086276.
+The authors report no conflict of interest related to this study.
+
+## Financial disclosures for the previous 12 months
+Dr. McKay has received research funding or support from the National Institutes of Health. Dr. McKay has received consulting fees from BioCircuit Technologies.
+
+# References

### 2019-12-20 README.Rmd
updated path locations
more changes
Added bibliographic information, initial frequency counts.

@@
     1. If there are files that need to be added to the remote, they can be placed in this folder or dragged / dropped into the GitHub interface.
-1. Path locations (used by Matlab code) are in `msBalance/paths.m`
+1. Path locations that need to be read by Matlab and R can be specified in `msBalance/paths.m`.
+1. Path locations might need to be modified in `msBalance/m/loadData.m`.
 1. The main analysis is in `msBalance/R/msBalance.Rmd`
@@
 1. A directory `csv/` is created to hold these files. The contents of this directory are not automatically loaded to GitHub.
+
+### Running the code
+
+1. Alter `paths.m` to point to the correct data directory.
+1. Run the Matlab code to load .mat files and populate patient-level data
+    * Run `m/loadData.m` to load source files and populate the `csv/` directory.
+    * A future version of this could call the Matlab script from R if useful.
+1. "knit" `R/msBalance.Rmd` to produce the study document.

### 2019-12-20 README.Rmd
updated docs

@@
 
-### Downloading from GitHub for development
+### What to read and edit
+1. This document explains how to run the code. It is an html version of `README.Rmd` that is "knit," which you can think of as sort-of interpreted or compiled.
+1. The document `R/msBalance.html` contains the output text and figures. It is the knit/interpreted version of `R/msBalance.Rmd`
 
+### How to run the code
+1. Alter `paths.m` in the main project directory to point to the correct data directory on your local machine.
+1. Run the Matlab code to load .mat files and populate patient-level data
+    * Run `m/loadData.m` to load source files and populate the `csv/` directory.
+    * A future version of this could call the Matlab script from R if useful.
+1. "knit" `R/msBalance.Rmd` to produce `R/msBalance.html`.
+
+### Notes related to the fact that there is a mixture of Matlab and R code
+1. The source files are in `.mat` format.
+1. There is a mixture of matlab preprocessing code (files in `m/`) into `.csv` files that R can then read (files in `R/`).
+1. A directory `csv/` is created to hold these files. The contents of this directory are not automatically loaded to GitHub.
+
+### How to download from GitHub for development
 1. Use RStudio to create a local copy of the project
@@
 1. Stage changed files, commit, and push changes with the git panel.
-
-### Navigating a mixture of Matlab and R code
-
-1. The source files are in `.mat` format.
-1. There is a mixture of matlab preprocessing code (files in `m/`) into `.csv` files that R can then read (files in `R/`).
-1. A directory `csv/` is created to hold these files. The contents of this directory are not automatically loaded to GitHub.
-
-### Running the code
-
-1. Alter `paths.m` to point to the correct data directory.
-1. Run the Matlab code to load .mat files and populate patient-level data
-    * Run `m/loadData.m` to load source files and populate the `csv/` directory.
-    * A future version of this could call the Matlab script from R if useful.
-1. "knit" `R/msBalance.Rmd` to produce the study document.

### 2020-01-15 README.Rmd
updated instructions on running Matlab commands.
updated matlab commands
updated msBalance.Rmd to call matlab preprocessing

@@
 ### How to run the code
-1. Alter `paths.m` in the main project directory to point to the correct data directory on your local machine.
-1. Run the Matlab code to load .mat files and populate patient-level data
-    * Run `m/loadData.m` to load source files and populate the `csv/` directory.
-    * A future version of this could call the Matlab script from R if useful.
-1. "knit" `R/msBalance.Rmd` to produce `R/msBalance.html`.
+1.
+Alter `paths.m` in the main project directory to point to the correct data directory on your local machine.
+This should contain the files `AllConsolidatedData_CS.mat` and `AllConsolidatedData_MS.mat`
+1. Also modify the `matlabCommand` string in paths.m. On Dr. McKay's machine, this string is `/Applications/MATLAB_R2019a.app/bin/matlab -nojvm -nodesktop -nosplash -r`
+1. "knit" `R/msBalance.Rmd` to produce `R/msBalance.html`. This will call matlab to create intermediary files if they are missing.
 

### 2020-01-15 paths.m

@@ -1,1 +1,2 @@
 dataDirectory = "/Users/lucasmckay/OneDrive - Emory University/2019 Peterson MS EMG/data/"
+matlabCommand = "/Applications/MATLAB_R2019a.app/bin/matlab -nojvm -nodesktop -nosplash -r"

### 2020-01-16 R/msBalance.Rmd
updated msBalance.Rmd to call matlab preprocessing
reorganized manuscript

@@
 ```{r}
+# set path variables
 source("../paths.m")
+
+# if csv files do not exist, run matlab batch steps
+if (!file.exists("../csv/c.csv")|!file.exists("../csv/m.csv")){
+  system(paste(matlabCommand, "'cd ../m; loadData; exit'"))
+}
 ```
@@
 
-# Methods and Results
+# Methods
 
 ## Data sources
+
+This study performed a secondary analysis on existing electromyographic recordings during balance responses in people with and without MS.
+Clinical and demographic characteristics of the study sample are summarized in Table 1.
+
 ```{r, message=FALSE, warning=FALSE}
@@
 
-This study performed a secondary analysis on existing electromyographic recordings during balance responses in people with and without MS.
+## Assessment protocol
+Complete details of the assessment protocol are avaialble in an earlier report.[@huisinga2014] 
+Briefly, subjects stood on a set of 2 custom hydraulic force platforms that translated forward together causing backward body sway and the activation of tibialis anterior (TA).
+subjects stood with arms folded across the chest, eyes open, with their feet at a fixed heel-to-heel distance of 10cm.
+Peak platform displacement was 4 cm, at a rate of 15 cm/s, which elicited feet-in-place balance corrections without stepping.
+Surface electromyography was recorded from the right TA of all subjects with two 2.5-cm^2^ surface electrodes placed approximately 2 cm apart, with a ground electrode on the lateral condyle.
+
+## Data processing
+Each individual EMG record was registered to the onset of platform motion and resampled.
+
 Individual subjects were as follows:

### 2020-01-16 README.Rmd
reorganized manuscript

--- 2019-12-20 README.Rmd
@@
 ---
+title: "Balance responses in MS"
+author: J.L. McKay
+date: "`r format(Sys.time(), '%d %B %Y')`"
 output: github_document
+always_allow_html: yes
 ---
@@
 ```{r setup, echo = FALSE}
-knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
+knitr::opts_chunk$set(echo = TRUE,
+                      collapse = TRUE,
+                      comment = "#>")
 ```
 
-# msBalance
-
-## Overview
-
-msBalance is an R project for analysis of balance responses in people with and without multiple sclerosis (MS).
+# Overview
+`msBalance` is an R project for analysis of balance responses in people with and without multiple sclerosis (MS).
 The source data were created by Jessie Huisinga, Ph.D., and collected by Daniel Peterson, Ph.D.
@@
 
-## Getting and installing msBalance
+# Use
 
-### Downloading from GitHub for development
+## What to read and edit
+1. The main document is `R/msBalance.Rmd`, which can be knit/interpreted into `R/msBalance.html`, which contains the main text and figures.
+1. This document explains how to run the code.
 
-1. Use RStudio to create a local copy of the project
-    1. In RStudio, start a new Project: `File > New Project > Version Control > Git`
-    1. In the “repository URL” paste: https://github.com/jlucasmckay/msBalance
-    1. On OS X, you can make this under `tmp` - this is a temporary file that can be deleted later
-    1. If there are files that need to be added to the remote, they can be placed in this folder or dragged / dropped into the GitHub interface.
-1. Path locations that need to be read by Matlab and R can be specified in `msBalance/paths.m`.
-1. Path locations might need to be modified in `msBalance/m/loadData.m`.
-1. The main analysis is in `msBalance/R/msBalance.Rmd`
-1. Stage changed files, commit, and push changes with the git panel.
+## How to run the code
+1. Alter `paths.m` in the main project directory so that: 
+    1. `dataDirectory` points at the correct data directory on your local machine. This should contain the files `AllConsolidatedData_CS.mat` and `AllConsolidatedData_MS.mat`.
+    1.
+`matlabCommand` points at the appropriate command-line version of Matlab for data preprocessing.
+On Dr.
+McKay's machine, this string is `/Applications/MATLAB_R2019a.app/bin/matlab -nojvm -nodesktop -nosplash -r`.
+1. "knit" `R/msBalance.Rmd` to produce `R/msBalance.html`. This will call matlab to create intermediary files if they are missing.
 
-### Navigating a mixture of Matlab and R code
-
+## Notes related to the fact that there is a mixture of Matlab and R code
 1. The source files are in `.mat` format.
@@
 
-### Running the code
-
-1. Alter `paths.m` to point to the correct data directory.
-1. Run the Matlab code to load .mat files and populate patient-level data
-    * Run `m/loadData.m` to load source files and populate the `csv/` directory.
-    * A future version of this could call the Matlab script from R if useful.
-1. "knit" `R/msBalance.Rmd` to produce the study document.
+## How to download from GitHub for development
+1. In RStudio, start a new Project: `File > New Project > Version Control > Git`.
+1. In the “repository URL” paste: https://github.com/jlucasmckay/msBalance.
+1. On OS X, you can make this under `tmp` - this is a temporary file that can be deleted later.
+1. If there are files that need to be added to the remote, they can be placed in this folder or dragged / dropped into the GitHub interface.
+1. Stage changed files, commit, and push changes with the git panel.

### 2020-01-16 R/msBalance.Rmd
got rid of amount of follow-up calculation

@@
 
-Individual subjects were as follows:
-
-```{r, warning=FALSE, message=FALSE}
-library(janitor)
-library(kableExtra)
-median0 = function (x) median(x[x>0], na.rm = TRUE)
-min0 = function (x) min(x[x>0], na.rm = TRUE)
-max0 = function (x) max(x[x>0], na.rm = TRUE)
-
-k = function (x) kable_styling(x, full_width = F, position = "left")
-t = followUp %>% tabyl(Group) %>% adorn_totals("row") %>% adorn_pct_formatting(digits = 0) %>% rename(N = n, `Percent`=percent)
-followUpAvailable = followUp %>% filter(Group=="MS") %>% ungroup() %>% mutate(f = `Follow-Up Available (Days)`>0) %>% summarize(fAvail = sum(f==TRUE,na.rm = TRUE)/n()) %>% pull() %>% scales::percent()
-followUpMed = median0(followUp$`Follow-Up Available (Days)`)
-followUpMin = min0(followUp$`Follow-Up Available (Days)`)
-followUpMax = max0(followUp$`Follow-Up Available (Days)`)
-kable(t) %>% k()
-```
-
-Of these, follow-up visits were available for several of the participants with MS (≈`r followUpAvailable`).
-Among the participants with MS and follow-up available, the median duration was `r followUpMed` days, and the range was from `r followUpMin` to `r followUpMax` days.
-The number of visits was distributed as follows:
-
-```{r, warning=FALSE, message=FALSE}
-library(janitor)
-library(kableExtra)
-t = followUp %>% tabyl(Group, `Study Visits Available (n)`)
-kable(t) %>% k() %>% add_header_above(c(" " = 1, "Number of Study Visits Available" = 4))
-```
-
 ## Exclusions

### 2020-01-16 R/msBalance.Rmd
Added first figure.

@@
 This study performed a secondary analysis on existing electromyographic recordings during balance responses in people with and without MS.
+Inclusion criteria for MS participants were as follows: diagnosis of MS by a MS neurologist and no relapses within 60 days.
+Inclusion criteria for all participants included ability to walk 25 ft (7.6 m) without an assistive device, and no other known biomechanical or neurologic conditions affecting stepping or balance.
 Clinical and demographic characteristics of the study sample are summarized in Table 1.
@@
 Complete details of the assessment protocol are avaialble in an earlier report.[@huisinga2014] 
-Briefly, subjects stood on a set of 2 custom hydraulic force platforms that translated forward together causing backward body sway and the activation of tibialis anterior (TA).
-subjects stood with arms folded across the chest, eyes open, with their feet at a fixed heel-to-heel distance of 10cm.
+Briefly, subjects stood on a set of 2 custom hydraulic force platforms that translated either forward or backward (Figure 1).
+Forward perturbations caused backward body sway and the activation of tibialis anterior (TA), and backward perturbations caused forward body sway, and the activation of medial gastrocnemius (MG).
+Subjects stood with arms folded across the chest, eyes open, with their feet at a fixed heel-to-heel distance of 10 cm.
 Peak platform displacement was 4 cm, at a rate of 15 cm/s, which elicited feet-in-place balance corrections without stepping.
 Surface electromyography was recorded from the right TA of all subjects with two 2.5-cm^2^ surface electrodes placed approximately 2 cm apart, with a ground electrode on the lateral condyle.
+
+```{r perturbationFigure, out.width = "25%", fig.cap = "Figure 1. Perturbation directions used in testing protocol.", fig.align="center"}
+knitr::include_graphics("figures/Perturbations.svg")
+```
 
@@
 
-# Disclosures
-
-## Funding sources and conflict of interest
+***
+### Funding sources and conflict of interest
 This study was supported in part by National Institutes of Health grant K25HD086276.
@@
 
-## Financial disclosures for the previous 12 months
+### Financial disclosures for the previous 12 months
 Dr. McKay has received research funding or support from the National Institutes of Health. Dr. McKay has received consulting fees from BioCircuit Technologies.
 
+***
 # References

### 2020-01-17 R/msBalance.Rmd
added first table 1

@@
   echo = FALSE,
+  message = FALSE,
+  warning = FALSE,
   collapse = TRUE,
@@
 ## Data sources
-
 This study performed a secondary analysis on existing electromyographic recordings during balance responses in people with and without MS.
@@
 
-```{r, message=FALSE, warning=FALSE}
+```{r loadData}
 library(vroom)
 library(dplyr)
+library(magrittr)
 library(stringr)
@@
 
+# consider only baseline data, and get rid of participants with missing minimal demographic data
+participants %<>% filter(!`Visit ID` %contains% "F") %>% filter(!is.na(Age)) %>% filter(Sex %in% c("M","F")) %>% filter(Age>0)
+
+# add a group condition - note that if_else in dplyr and ifelse in base r are different!
+participants %<>% mutate(Group = if_else(Participant %contains% "MSC", "Control", "MS")) %>%
+  mutate(Group = relevel(factor(Group), "MS")) %>% select(Group, Participant, Age, Sex, everything())
+```
+
+```{r layout}
+library(captioner)
+tableNumbers <- captioner(prefix = "Table")
+captionDemographics = tableNumbers(name = "demographics", caption = "Clinical and demographic features of the study population.")
+figureNumbers <- captioner(prefix = "Figure")
+captionCartoon = figureNumbers(name = "cartoon", caption = "Perturbation directions used in testing protocol.")
 ```
@@
 Complete details of the assessment protocol are avaialble in an earlier report.[@huisinga2014] 
-Briefly, subjects stood on a set of 2 custom hydraulic force platforms that translated either forward or backward (Figure 1).
-Forward perturbations caused backward body sway and the activation of tibialis anterior (TA), and backward perturbations caused forward body sway, and the activation of medial gastrocnemius (MG).
+Briefly, subjects stood on a set of 2 custom hydraulic force platforms that translated either forward or backward (`r figureNumbers("cartoon", display = "cite")`).
+In this study, "perturbation direction" refers to the direction of postural sway caused by the surface translation.
+Therefore, a backward perturbation refers to forward movement of the support surface causing backward body sway and the activation of tibialis anterior (TA),
+whereas forward perturbations refer to backward translation of the support surface causing forward body sway and the activation of medial gastrocnemius (MG).
 Subjects stood with arms folded across the chest, eyes open, with their feet at a fixed heel-to-heel distance of 10 cm.
@@
 
-```{r perturbationFigure, out.width = "25%", fig.cap = "Figure 1. Perturbation directions used in testing protocol.", fig.align="center"}
+```{r perturbationFigure, out.width = "25%", fig.align = "center", fig.cap = captionCartoon}
 knitr::include_graphics("figures/Perturbations.svg")
@@
 
-## Exclusions
-Beginning with N=TK records, TK were excluded (N=TK, MS, N=TK, Control) due to incomplete demographic information.
-Characteristics of the remaining participants included in analyses are summarized in Table 1.
+## Demographic and clinical characteristics
+Demographic and clinical characteristics of the study population are summarized in `r tableNumbers("demographics", display = "cite")`.
 
-## Demographic and Clinical Characteristics
+```{r table1}
+library(janitor)
+library(kableExtra)
+participants %>% tabyl(Sex, Group) %>% mutate(Total = MS + Control) %>% 
+  kable() %>% kable_styling(full_width = F, position = "left")
+```
 

### 2020-01-17 R/msBalance.Rmd
updated table and figure captions

@@
 # Methods
-
-## Data sources
-This study performed a secondary analysis on existing electromyographic recordings during balance responses in people with and without MS.
-Inclusion criteria for MS participants were as follows: diagnosis of MS by a MS neurologist and no relapses within 60 days.
-Inclusion criteria for all participants included ability to walk 25 ft (7.6 m) without an assistive device, and no other known biomechanical or neurologic conditions affecting stepping or balance.
-Clinical and demographic characteristics of the study sample are summarized in Table 1.
 
@@
 ```{r layout}
+# layout tables and figures
 library(captioner)
@@
 captionCartoon = figureNumbers(name = "cartoon", caption = "Perturbation directions used in testing protocol.")
+```
+
+## Data sources
+This study analyzed existing electromyographic recordings during balance responses in people with and without MS.
+An earlier analysis of a subset of these records is available.[@peterson2016]
+Inclusion criteria for MS participants were as follows: diagnosis of MS by a MS neurologist and no relapses within 60 days.
+Inclusion criteria for all participants included ability to walk 25 ft (7.6 m) without an assistive device, and no other known biomechanical or neurologic conditions affecting stepping or balance.
+Clinical and demographic characteristics of the study sample are summarized in `r tableNumbers("demographics", display = "cite")`.
+
+```{r demographics}
+library(janitor)
+library(kableExtra)
+participants %>% tabyl(Sex, Group) %>% mutate(Total = MS + Control) %>% 
+  kable(caption = captionDemographics) %>% kable_styling(full_width = F, position = "left")
 ```
@@
 
-```{r perturbationFigure, out.width = "25%", fig.align = "center", fig.cap = captionCartoon}
+```{r perturbationFigure, out.width = "25%", fig.align = "left", fig.cap = captionCartoon}
 knitr::include_graphics("figures/Perturbations.svg")
@@
 Demographic and clinical characteristics of the study population are summarized in `r tableNumbers("demographics", display = "cite")`.
-
-```{r table1}
-library(janitor)
-library(kableExtra)
-participants %>% tabyl(Sex, Group) %>% mutate(Total = MS + Control) %>% 
-  kable() %>% kable_styling(full_width = F, position = "left")
-```
 

### 2020-01-28 R/msBalance.Rmd
first interpolation of EMG data

@@
 
-# if csv files do not exist, run matlab batch steps
+# if csv files do not exist, run matlab batch steps to populate them
 if (!file.exists("../csv/c.csv")|!file.exists("../csv/m.csv")){
@@
 ## Data processing
-Each individual EMG record was registered to the onset of platform motion and resampled.
+Each individual EMG record was registered to the onset of platform motion and resampled to a common timebase
 

### 2020-01-28 m/interpolateRawData.m

@@ -1,0 +1,73 @@
+function out = interpolateRawData(in, fs, ts)
+
+% set output frequency to default if not specified
+if nargin<2
+    fs = 1000;
+end
+if nargin<3
+    ts = [-0.5:(1/fs):(2.0-(1/fs))];
+end
+
+nR = size(in,1);
+nC = length(ts);
+
+
+[e1, e2, e3, e4, p] = deal(nan(nR, nC));
+
+DataError = false(nR, 1);
+Participant = string(in.Subj);
+Trial = string(in.trial);
+try
+    Date = in.obsDate;
+catch
+    Date = NaT(nR,1);
+end
+
+for t = 1:nR
+    emgTime = in.time{t,:};
+    emg1 = in.emg1{t,:};
+    emg2 = in.emg1{t,:};
+    emg3 = in.emg1{t,:};
+    emg4 = in.emg1{t,:};
+    treadmillPosition = in.Treadmill_marker_position{t,:};
+    treadmillTime = in.Treadmill_Marker_Time{t,:};
+    onsetSample = identifyOnsetSample(treadmillPosition);
+    onsetTime = treadmillTime(onsetSample);
+    
+    % identify sample frequencies
+    sampleFrequencyEmg = 1/(emgTime(2)-emgTime(1));
+    sampleFrequencyTreadmill = 1/(treadmillTime(2)-treadmillTime(1));
+
+    % resample signals
+    try
+    [emg1Res, emg1ResTime] = resample(emg1, emgTime - onsetTime, fs);
+    [emg2Res, emg2ResTime] = resample(emg2, emgTime - onsetTime, fs);
+    [emg3Res, emg3ResTime] = resample(emg3, emgTime - onsetTime, fs);
+    [emg4Res, emg4ResTime] = resample(emg4, emgTime - onsetTime, fs);
+    [treadmillPositionRes, treadmillPositionResTime] = resample(treadmillPosition, treadmillTime - onsetTime, fs);
+    
+    % interpolate to common timebase and save
+    e1(t,:) = interp1(emg1ResTime, emg1Res, ts, 'linear', 0)';
+    e2(t,:) = interp1(emg2ResTime, emg2Res, ts, 'linear', 0)';
+    e3(t,:) = interp1(emg3ResTime, emg3Res, ts, 'linear', 0)';
+    e4(t,:) = interp1(emg4ResTime, emg4Res, ts, 'linear', 0)';
[…]

### 2020-01-28 m/inversewavelettransform.m

@@ -1,0 +1,9 @@
+function data = inversewavelettransform(wavedata,waveparams)
+dwtmode(waveparams.wavemode);
+% Get the length of the data records
+[ntrls,npnts] = size(wavedata);
+data = zeros(size(wavedata));
+for i = 1:ntrls
+    data(i,:) = waverec(wavedata(i,:),waveparams.wL,waveparams.wavestr);
+end
+end

### 2020-01-28 m/preprocessData.m

@@ -1,0 +1,47 @@
+% /Applications/MATLAB_R2019a.app/bin/matlab -nojvm -nodesktop -nosplash
+% preprocessData.m
+% 2020 01 27
+% JLM
+
+% free up memory
+clear
+close all
+
+% specify data directory
+run(".."+filesep+"paths.m")
+
+% load data for CS and MS subjects, which are organized as tables
+load(dataDirectory + "AllConsolidatedData_CS.mat")
+c = T;
+load(dataDirectory + "AllConsolidatedData_MS.mat")
+m = T;
+clear T
+
+% set a flag for interactive use, making plots, etc.
+interactive = 1;
+
+% interpolate data to common timebase
+m2 = interpolateRawData(m);
+c2 = interpolateRawData(c);
+
+% concatenate data into common table in RAM
+d = [m2; c2];
+
+% wavelet transform
+
+% do statistics
+
+% inverse transform
+
+
+% load data dictionary
+vars = readtable(dataDirectory + "dataDictionary.csv")
+idVars = vars.name(vars.class == "id" | vars.class == "dem")
+timeVars = vars.name(vars.class == "time")
+
+% pipe the patient-level-data to two csv files that can be combined in R
+writetable(m(:,ismember(m.Properties.VariableNames,idVars)), csvDirectory + "m.csv")
+writetable(c(:,ismember(c.Properties.VariableNames,idVars)), csvDirectory + "c.csv")
+
+% free up memory
+clear

### 2020-01-28 m/wavelettransform.m

@@ -1,0 +1,22 @@
+function [wavedata, waveparams] = wavelettransform(data)
+% Get the length of the data records
+[ntrls,npnts] = size(data);
+% Wavelet to use: 3-rd order Coiflet.
+wavestr = 'coif3';
+% Extension mode: periodic, so that what is returned is the same length as
+% the data.
+wavemode = 'per';
+dwtmode(wavemode);
+% Find the maximum possible level of the wavelet decomposition:
+lev = wmaxlev([1 npnts],wavestr);
+% Loop through and transform
+wavedata = zeros(size(data));
+for i = 1:ntrls
+    x = data(i,:);
+    [wx,wL] = wavedec(x,lev,wavestr);
+    wavedata(i,:) = wx;
+end
+waveparams.wL = wL;
+waveparams.wavestr = wavestr;
+waveparams.wavemode = wavemode;
+end

### 2020-01-28 m/interpolateRawData.m
matlab updates to interpolate data

@@ -1,4 +1,4 @@
-function out = interpolateRawData(in, fs, ts)
+function [out, ts] = interpolateRawData(in, fs, ts)
 
 % set output frequency to default if not specified
 if nargin<2

### 2020-01-28 m/preprocessData.m

@@ -17,15 +17,30 @@ load(dataDirectory + "AllConsolidatedData_MS.mat")
 m = T;
 clear T
 
-% set a flag for interactive use, making plots, etc.
-interactive = 1;
-
 % interpolate data to common timebase
-m2 = interpolateRawData(m);
-c2 = interpolateRawData(c);
+[m2, ~] = interpolateRawData(m);
+[c2, Time] = interpolateRawData(c);
 
-% concatenate data into common table in RAM
+% concatenate data into common table
 d = [m2; c2];
+
+% during development, there were a small number of trials (4) with loading
+% errors. exclude these.
+d(d.DataError,:) = [];
+d.DataError = [];
+
+% explode variables
+Participant = d.Participant;
+Date = d.Date;
+Trial = d.Trial;
+Emg1 = d.e1;
+Emg2 = d.e2;
+Emg3 = d.e3;
+Emg4 = d.e4;
+Treadmill = d.p;
+
+% save into temporary directory
+7+3
 
 % wavelet transform
 

### 2020-01-28 m/preprocessData.m
added csv output

@@ -10,6 +10,9 @@ close all
 % specify data directory
 run(".."+filesep+"paths.m")
 
+% specify temporary directory
+tmpDir = ".."+filesep+"tmp"+filesep
+
 % load data for CS and MS subjects, which are organized as tables
 load(dataDirectory + "AllConsolidatedData_CS.mat")
 c = T;
@@ -39,7 +42,12 @@ Emg3 = d.e3;
 Emg4 = d.e4;
 Treadmill = d.p;
 
+% filter EMG
+
+% normalize EMG
+
 % save into temporary directory
+writematrix(Treadmill, tmpDir+"Treadmill.csv")
 7+3
 
 % wavelet transform

### 2020-01-28 R/msBalance.Rmd
added call to preprocessData.m

@@
 ## Data processing
-Each individual EMG record was registered to the onset of platform motion and resampled to a common timebase
+Each individual EMG record was registered to the onset of treadmill motion and resampled to a common timebase.
+
+```{r}
+# if trial-level data files do not exist, run matlab batch steps to populate them
+if (!file.exists("../tmp/Treadmill.csv")){
+  system(paste(matlabCommand, "'cd ../m; preprocessData; exit'"))
+}
+```
 

### 2020-01-28 m/loadData.m

@@ -26,6 +26,3 @@ timeVars = vars.name(vars.class == "time")
 % pipe the patient-level-data to two csv files that can be combined in R
 writetable(m(:,ismember(m.Properties.VariableNames,idVars)), csvDirectory + "m.csv")
 writetable(c(:,ismember(c.Properties.VariableNames,idVars)), csvDirectory + "c.csv")
-
-% free up memory
-clear

### 2020-01-28 m/preprocessData.m

@@ -8,10 +8,10 @@ clear
 close all
 
 % specify data directory
-run(".."+filesep+"paths.m")
+run(".." + filesep + "paths.m")
 
 % specify temporary directory
-tmpDir = ".."+filesep+"tmp"+filesep
+tmpDir = ".." + filesep + "tmp" + filesep
 
 % load data for CS and MS subjects, which are organized as tables
 load(dataDirectory + "AllConsolidatedData_CS.mat")
@@ -47,24 +47,10 @@ Treadmill = d.p;
 % normalize EMG
 
 % save into temporary directory
-writematrix(Treadmill, tmpDir+"Treadmill.csv")
-7+3
+writematrix(Treadmill, tmpDir + "Treadmill.csv")
 
 % wavelet transform
 
 % do statistics
 
 % inverse transform
-
-
-% load data dictionary
-vars = readtable(dataDirectory + "dataDictionary.csv")
-idVars = vars.name(vars.class == "id" | vars.class == "dem")
-timeVars = vars.name(vars.class == "time")
-
-% pipe the patient-level-data to two csv files that can be combined in R
-writetable(m(:,ismember(m.Properties.VariableNames,idVars)), csvDirectory + "m.csv")
-writetable(c(:,ismember(c.Properties.VariableNames,idVars)), csvDirectory + "c.csv")
-
-% free up memory
-clear

### 2020-01-28 R/msBalance.Rmd
changed csv directory to tmp/

@@
 # if csv files do not exist, run matlab batch steps to populate them
-if (!file.exists("../csv/c.csv")|!file.exists("../csv/m.csv")){
+if (!file.exists("../csv/c.csv")|!file.exists("../tmp/m.csv")){
   system(paste(matlabCommand, "'cd ../m; loadData; exit'"))
@@
 # load participant-level information
-m = vroom("../csv/m.csv") %>% mutate(Group = "MS") %>% mutate(Age = round(age)) %>% mutate(Sex = toupper(gender)) %>% rename(Participant = Subj)
-c = vroom("../csv/c.csv") %>% mutate(Group = "Control") %>% mutate(Age = round(age)) %>% mutate(Sex = toupper(gender)) %>% rename(Participant = Subj)
+m = vroom("../tmp/m.csv") %>% mutate(Group = "MS") %>% mutate(Age = round(age)) %>% mutate(Sex = toupper(gender)) %>% rename(Participant = Subj)
+c = vroom("../tmp/c.csv") %>% mutate(Group = "Control") %>% mutate(Age = round(age)) %>% mutate(Sex = toupper(gender)) %>% rename(Participant = Subj)
 

### 2020-01-28 m/loadData.m

@@ -7,10 +7,10 @@
 clear
 
 % specify data directory
-run(".."+filesep+"paths.m")
+run(".." + filesep + "paths.m")
 
-% specify csv file directory
-csvDirectory = ".."+filesep+"csv"+filesep
+% specify temporary directory
+csvDirectory = ".." + filesep + "tmp" + filesep
 
 % load data for CS and MS subjects, which are organized as tables
 load(dataDirectory + "AllConsolidatedData_CS.mat")

### 2020-01-28 R/msBalance.Rmd
added processing code to time-align emg and kinematic signals. discovered some significant issues with kinematic marker dropout, sent query to dan about perturbationOnset data.

@@
 ## Data processing
-Each individual EMG record was registered to the onset of treadmill motion and resampled to a common timebase.
+Each individual EMG record was registered to the onset of treadmill motion and resampled to a common sample rate (1 kHz), and processed (high-pass, 35 Hz, de- mean, rectify, low-pass, 40 Hz) as in previous studies.[@lang2019; @mckay2016]
 
@@
 # if trial-level data files do not exist, run matlab batch steps to populate them
-if (!file.exists("../tmp/Treadmill.csv")){
+if (!file.exists("../tmp/Participant.csv")|
+    !file.exists("../tmp/Date.csv")|
+    !file.exists("../tmp/Trial.csv")|
+    !file.exists("../tmp/Time.csv")|
+    !file.exists("../tmp/Treadmill.csv")|
+    !file.exists("../tmp/Emg1.csv")|
+    !file.exists("../tmp/Emg2.csv")|
+    !file.exists("../tmp/Emg3.csv")|
+    !file.exists("../tmp/Emg4.csv")){
   system(paste(matlabCommand, "'cd ../m; preprocessData; exit'"))
 }
+trialData = cbind(
+vroom("../tmp/Participant.csv", col_names = FALSE) %>% select(Participant = X1),
+vroom("../tmp/Date.csv", col_names = FALSE) %>% select(Date = X1),
+vroom("../tmp/Trial.csv", col_names = FALSE) %>% select(Trial = X1)
+) %>% as_tibble() %>% mutate(Date = ymd(Date))
+
+time = vroom("../tmp/Time.csv", col_names = FALSE)
+treadmill = vroom("../tmp/Treadmill.csv", col_names = FALSE)
+emg1 = vroom("../tmp/Emg1.csv", col_names = FALSE)
+emg2 = vroom("../tmp/Emg2.csv", col_names = FALSE)
+emg3 = vroom("../tmp/Emg3.csv", col_names = FALSE)
+emg4 = vroom("../tmp/Emg4.csv", col_names = FALSE)
+
 ```

### 2020-01-28 m/preprocessData.m

@@ -20,9 +20,12 @@ load(dataDirectory + "AllConsolidatedData_MS.mat")
 m = T;
 clear T
 
+% set default sample frequency
+fs = 1000;
+
 % interpolate data to common timebase
-[m2, ~] = interpolateRawData(m);
-[c2, Time] = interpolateRawData(c);
+[m2, ~] = interpolateRawData(m, fs);
+[c2, Time] = interpolateRawData(c, fs);
 
 % concatenate data into common table
 d = [m2; c2];
@@ -32,22 +35,26 @@ d = [m2; c2];
 d(d.DataError,:) = [];
 d.DataError = [];
 
-% explode variables
+% explode variables and pre-Process
 Participant = d.Participant;
-Date = d.Date;
+Date = datestr(d.Date,'yyyy-mm-dd');
 Trial = d.Trial;
-Emg1 = d.e1;
-Emg2 = d.e2;
-Emg3 = d.e3;
-Emg4 = d.e4;
-Treadmill = d.p;
-
-% filter EMG
-
-% normalize EMG
+Emg1 = preprocessEmg(d.e1, fs);
+Emg2 = preprocessEmg(d.e2, fs);
+Emg3 = preprocessEmg(d.e3, fs);
+Emg4 = preprocessEmg(d.e4, fs);
+Treadmill = preprocessTreadmill(d.p);
 
 % save into temporary directory
-writematrix(Treadmill, tmpDir + "Treadmill.csv")
+writematrix(Participant, tmpDir + "Participant.csv");
+writematrix(Date, tmpDir + "Date.csv");
+writematrix(Trial, tmpDir + "Trial.csv");
+writematrix(Time, tmpDir + "Time.csv");
+writematrix(Treadmill, tmpDir + "Treadmill.csv");
+writematrix(Emg1, tmpDir + "Emg1.csv");
+writematrix(Emg2, tmpDir + "Emg2.csv");
+writematrix(Emg3, tmpDir + "Emg3.csv");
+writematrix(Emg4, tmpDir + "Emg4.csv");
 
 % wavelet transform
 

### 2020-01-28 m/preprocessEmg.m

@@ -1,0 +1,54 @@
+function emg = preprocessEmg(emg, fs)
+emgOut = nan(size(emg));
+
+for i = 1:size(emg,1)
+    in = emg(i,:)';
+    out = proc_process_emg(in, fs);
+    emgOut(i,:) = out';
+end
+end
+
+function emg = proc_process_emg(EMG, SampleRate)
+% ver. 2 - March 10, 2010 (Jeff Bingham) Switched versioning to mcollapse
+% ver. 1 - February 6, 2010 (Jeff Bingham) File created from portions of     
+%          the previous 'proc2axis (August 28, 2009)' that were involved
+%          with processing EMG data from C3D files.
+%
+% emg = proc_process_emg(EMG, SampleRate)
+%
+%            Function that filters and rectifies raw EMG data
+%
+% #########################################################################
+%
+%
+% SUPPORTING FUNCTIONS:
+%
+% ARGUMENTS:
+%  EMG        [m x n] ... The "m" samples for "n" different EMG channels of
+%                         raw unprocessed data
+%  SampleRate   INT ..... The rate of samples per second the EMG data was
+%                         collected at
+%
+% OUTPUTS:
+%  emg        [m x n] ... The "m" samples of "n" filtered EMG channels
+%
+% #########################################################################
+
+nyquist_frequency = SampleRate/2;
+
+%%% Create filters
+% High pass filter at 35 Hz
+[filt_high_B,filt_high_A] = butter(3,35/nyquist_frequency,'high');
+% Low pass filter at 40 Hz
+[filt_low_B,filt_low_A] = butter(3,40/nyquist_frequency,'low');
+
+%%% Filter EMG signals
+% High pass filter at 35 Hz
+emg = filtfilt(filt_high_B, filt_high_A,EMG);
+% Demean and rectify
+emg_mean = repmat(mean(emg(1:SampleRate,:),1),size(emg,1),1);
+emg = abs(emg-emg_mean);
+% Low pass filter at 40 Hz
+emg = filtfilt(filt_low_B, filt_low_A,emg);
+
+end

### 2020-01-28 m/preprocessTreadmill.m

@@ -1,0 +1,5 @@
+function Treadmill = preprocessTreadmill(Treadmill)
+nSamp = 50;
+ZeroVal = nanmean(Treadmill(:,1:nSamp),2);
+Treadmill = Treadmill - repmat(ZeroVal, 1, size(Treadmill,2));
+end

### 2020-01-29 m/interpolateRawData.m
tiny changes

@@ -11,7 +11,6 @@ end
 nR = size(in,1);
 nC = length(ts);
 
-
 [e1, e2, e3, e4, p] = deal(nan(nR, nC));
 
 DataError = false(nR, 1);

### 2020-01-29 m/interpolateRawData.m
altered matlab to use "perturbationOnset" field for onset time

@@ -30,8 +30,10 @@ for t = 1:nR
     emg4 = in.emg1{t,:};
     treadmillPosition = in.Treadmill_marker_position{t,:};
     treadmillTime = in.Treadmill_Marker_Time{t,:};
-    onsetSample = identifyOnsetSample(treadmillPosition);
-    onsetTime = treadmillTime(onsetSample);
+%     onsetSample = identifyOnsetSample(treadmillPosition);
+%     onsetTime = treadmillTime(onsetSample);
+    onsetSample = round(in.perturbationOnset(t));
+    onsetTime = in.time{t,onsetSample};
     
     % identify sample frequencies
     sampleFrequencyEmg = 1/(emgTime(2)-emgTime(1));

### 2020-01-29 m/preprocessEmg.m
changes to correct preprocessEmg.m to return the processed rather than original variable

@@ -6,6 +6,7 @@ for i = 1:size(emg,1)
     out = proc_process_emg(in, fs);
     emgOut(i,:) = out';
 end
+emg = emgOut;
 end
 
 function emg = proc_process_emg(EMG, SampleRate)
@@ -45,9 +46,13 @@ nyquist_frequency = SampleRate/2;
 %%% Filter EMG signals
 % High pass filter at 35 Hz
 emg = filtfilt(filt_high_B, filt_high_A,EMG);
-% Demean and rectify
+
+% de-mean
 emg_mean = repmat(mean(emg(1:SampleRate,:),1),size(emg,1),1);
+
+% rectify
 emg = abs(emg-emg_mean);
+
 % Low pass filter at 40 Hz
 emg = filtfilt(filt_low_B, filt_low_A,emg);
 

### 2020-01-29 R/msBalance.Rmd
added material to make plots and pull into R

@@
 whereas forward perturbations refer to backward translation of the support surface causing forward body sway and the activation of medial gastrocnemius (MG).
-Subjects stood with arms folded across the chest, eyes open, with their feet at a fixed heel-to-heel distance of 10 cm.
+
+During each asssessment, subjects stood with arms folded across the chest, eyes open, with their feet at a fixed heel-to-heel distance of 10 cm.
+Six perturbations were delivered, three forward, followed by three backward.
 Peak platform displacement was 4 cm, at a rate of 15 cm/s, which elicited feet-in-place balance corrections without stepping.
-Surface electromyography was recorded from the right TA of all subjects with two 2.5-cm^2^ surface electrodes placed approximately 2 cm apart, with a ground electrode on the lateral condyle.
+Surface electromyography was recorded with pairs 2.5-cm^2^ surface electrodes placed approximately 2 cm apart, with a ground electrode on the lateral condyle. 
+For this preliminary analysis, we considered only the activity of tibialis anterior of the right leg during forward perturbations.
 
@@
 
+# the treadmill records are not very useful, as they have not been cleaned up manually, so let's plan to assume Dan's coding of the study conditions.
+# To your questions
+#  
+# 1)      We have bilateral TA and medial Gastroc. Specifically
+# a.       Ch. 1= Rt. TA
+# b.       Ch. 2= Lf. TA
+# c.       Ch. 3= Rt. MG
+# d.       Ch. 4= Lf. MG
+# 2)      We have 6 trials per participant per visit (some people have multiple visits). Within each visit, they complete 6 trials…
+# a.       First 3 are FORWARD steps / loss of balance (i.e. the treadmill belt moves backward underfoot)
+# b.       Next 3 are BACKWARD steps / loss of balance (i.e. the treadmill belt moves forward underfoot)
+
+# the dataset emg1 corresponds to right TA. reconfigure it to enable tidy plotting.
+
 ```

### 2020-01-29 m/interpolateRawData.m

@@ -25,15 +25,15 @@ end
 for t = 1:nR
     emgTime = in.time{t,:};
     emg1 = in.emg1{t,:};
-    emg2 = in.emg1{t,:};
-    emg3 = in.emg1{t,:};
-    emg4 = in.emg1{t,:};
+    emg2 = in.emg2{t,:};
+    emg3 = in.emg3{t,:};
+    emg4 = in.emg4{t,:};
     treadmillPosition = in.Treadmill_marker_position{t,:};
     treadmillTime = in.Treadmill_Marker_Time{t,:};
-%     onsetSample = identifyOnsetSample(treadmillPosition);
-%     onsetTime = treadmillTime(onsetSample);
+    
+    % use supplied onset time, round to nearest sample
     onsetSample = round(in.perturbationOnset(t));
-    onsetTime = in.time{t,onsetSample};
+    onsetTime = in.time{t}(onsetSample);
     
     % identify sample frequencies
     sampleFrequencyEmg = 1/(emgTime(2)-emgTime(1));

### 2020-01-29 m/preprocessData.m

@@ -7,6 +7,9 @@
 clear
 close all
 
+% set flag to run interactive and make plots
+runInteractive = true;
+
 % specify data directory
 run(".." + filesep + "paths.m")
 
@@ -44,6 +47,22 @@ Emg2 = preprocessEmg(d.e2, fs);
 Emg3 = preprocessEmg(d.e3, fs);
 Emg4 = preprocessEmg(d.e4, fs);
 Treadmill = preprocessTreadmill(d.p);
+
+if runInteractive
+    f = figure;
+    nr = 5;
+    nc = 1;
+    subplot(nr,nc,1)
+    plot(Time,Treadmill)
+    subplot(nr,nc,2)
+    plot(Time,Emg1)    
+    subplot(nr,nc,3)
+    plot(Time,Emg2)    
+    subplot(nr,nc,4)
+    plot(Time,Emg3)    
+    subplot(nr,nc,5)
+    plot(Time,Emg4)    
+end
 
 % save into temporary directory
 writematrix(Participant, tmpDir + "Participant.csv");

### 2020-01-29 m/preprocessData.m
transposed matrices in matlab so that R likes them better

@@ -8,7 +8,7 @@ clear
 close all
 
 % set flag to run interactive and make plots
-runInteractive = true;
+runInteractive = false;
 
 % specify data directory
 run(".." + filesep + "paths.m")
@@ -64,16 +64,17 @@ if runInteractive
     plot(Time,Emg4)    
 end
 
-% save into temporary directory
+% save into temporary directory. note that timeseries are much easier to
+% work with with time samples going down the rows, so transpose these.
 writematrix(Participant, tmpDir + "Participant.csv");
 writematrix(Date, tmpDir + "Date.csv");
 writematrix(Trial, tmpDir + "Trial.csv");
-writematrix(Time, tmpDir + "Time.csv");
-writematrix(Treadmill, tmpDir + "Treadmill.csv");
-writematrix(Emg1, tmpDir + "Emg1.csv");
-writematrix(Emg2, tmpDir + "Emg2.csv");
-writematrix(Emg3, tmpDir + "Emg3.csv");
-writematrix(Emg4, tmpDir + "Emg4.csv");
+writematrix(Time', tmpDir + "Time.csv");
+writematrix(Treadmill', tmpDir + "Treadmill.csv");
+writematrix(Emg1', tmpDir + "Emg1.csv");
+writematrix(Emg2', tmpDir + "Emg2.csv");
+writematrix(Emg3', tmpDir + "Emg3.csv");
+writematrix(Emg4', tmpDir + "Emg4.csv");
 
 % wavelet transform
 

### 2020-01-30 R/msBalance.Rmd
first implementation of plotting timeseries data in R

@@
 
-# identify enrollment date for each participant and calculate amount of observation time
+# identify enrollment date for each participant and calculate amount of time
 participants = participants %>% left_join(participants %>% group_by(Participant) %>% summarize(`Enrollment Date` = min(`Visit Date`, na.rm = TRUE))) %>% 
@@
 Clinical and demographic characteristics of the study sample are summarized in `r tableNumbers("demographics", display = "cite")`.
-
-```{r demographics}
-library(janitor)
-library(kableExtra)
-participants %>% tabyl(Sex, Group) %>% mutate(Total = MS + Control) %>% 
-  kable(caption = captionDemographics) %>% kable_styling(full_width = F, position = "left")
-```
 
@@
 }
+
+# code the trial-level information so that each is coded.
+# this call also codes the groups and isolates first visits based on the subject code.
+# finally, this code isolates the trial number and direction from the Trial code, _t1-_t6.
 trialData = cbind(
@@
   vroom("../tmp/Trial.csv", col_names = FALSE) %>% select(Trial = X1)
-) %>% as_tibble() %>% mutate(Date = ymd(Date))
+) %>% as_tibble() %>%
+  mutate(Date = ymd(Date)) %>%
+  mutate(`Code` = paste0("X", row_number())) %>% 
+  mutate(`Study Group` = if_else(Participant %contains% "MSC", "Control", "MS")) %>% 
+  mutate(Visit = if_else(Participant %contains% "F", "Follow-up", "Initial")) %>% 
+  mutate(`Trial Number` = Trial %>% str_extract("_t\\d") %>% str_extract("\\d") %>% as.numeric()) %>% 
+  mutate(`Perturbation Direction` = if_else(`Trial Number` <= 3, "Forward", "Backward"))
 
-time = vroom("../tmp/Time.csv", col_names = FALSE)
+time = vroom("../tmp/Time.csv", col_names = FALSE) %>% select(Time = X1)
 treadmill = vroom("../tmp/Treadmill.csv", col_names = FALSE)
@@
 # the dataset emg1 corresponds to right TA. reconfigure it to enable tidy plotting.
+taR = time %>%  cbind(emg1) %>% as_tibble() %>% tidyr::pivot_longer(-Time, names_to = "Code", values_to = "V") %>%
+  mutate(Muscle = "Tibialis Anterior") %>% mutate(Side = "Right") %>% select(`Code`, Muscle, Side, everything())
 
+# I cannot figure out how to get dplyr::left_join to join by a variable with backticks, so rename temporarily
+taR = left_join(taR, trialData, by = "Code") %>% 
+  select(`Study Group`,Participant, Date, Trial, `Code`, `Trial Number`, `Perturbation Direction`, Muscle, Side, everything()) %>% 
+  arrange(`Study Group`, Participant, `Trial Number`, Time)
+
+# note that a large amount of the data (40%) is follow-up, with different visit amounts. for now, let's just consider the initial visit.
+taR = taR %>% filter(Visit == "Initial")
+
+# consider only those participants for whom both some EMG and minimal demographic data are available.
+includedParticipants = intersect(taR$Participant, participants$Participant)
+taR = taR %>% filter(Participant %in% includedParticipants)
+participants = participants %>% filter(Participant %in% includedParticipants)
+
+```
+
+```{r demographics}
+library(janitor)
+library(kableExtra)
+participants %>% tabyl(Sex, Group) %>% mutate(Total = MS + Control) %>% 
+  kable(caption = captionDemographics) %>% kable_styling(full_width = F, position = "left")
+```
+
+```{r rawEmg}
+library(ggplot2)
+p = ggplot(taR, aes(x = Time, y = V, group = Code)) + geom_line() + facet_grid(`Study Group` ~ `Perturbation Direction`)
+p
 ```

### 2020-01-30 R/msBalance.Rmd
updated figure to be backward - forward

@@
 participants %<>% mutate(Group = if_else(Participant %contains% "MSC", "Control", "MS")) %>%
-  mutate(Group = relevel(factor(Group), "MS")) %>% select(Group, Participant, Age, Sex, everything())
+  mutate(Group = relevel(factor(Group), "Control")) %>% select(Group, Participant, Age, Sex, everything())
 ```
@@
 captionCartoon = figureNumbers(name = "cartoon", caption = "Perturbation directions used in testing protocol.")
+captionRawEmg = figureNumbers(name = "rawEmg", caption = "EMG traces stratified by perturbation direction and study group.")
 ```
@@
 Inclusion criteria for all participants included ability to walk 25 ft (7.6 m) without an assistive device, and no other known biomechanical or neurologic conditions affecting stepping or balance.
-Clinical and demographic characteristics of the study sample are summarized in `r tableNumbers("demographics", display = "cite")`.
 
@@
 Complete details of the assessment protocol are avaialble in an earlier report.[@huisinga2014] 
-Briefly, subjects stood on a set of 2 custom hydraulic force platforms that translated either forward or backward (`r figureNumbers("cartoon", display = "cite")`).
+During each asssessment, subjects stood with arms folded across the chest, eyes open, with their feet at a fixed heel-to-heel distance of 10 cm on a set of 2 custom hydraulic force platforms that translated either forward or backward (`r figureNumbers("cartoon", display = "cite")`).
 In this study, "perturbation direction" refers to the direction of postural sway caused by the surface translation.
@@
 whereas forward perturbations refer to backward translation of the support surface causing forward body sway and the activation of medial gastrocnemius (MG).
-
-During each asssessment, subjects stood with arms folded across the chest, eyes open, with their feet at a fixed heel-to-heel distance of 10 cm.
 Six perturbations were delivered, three forward, followed by three backward.
 Peak platform displacement was 4 cm, at a rate of 15 cm/s, which elicited feet-in-place balance corrections without stepping.
+
+## Electromyography
 Surface electromyography was recorded with pairs 2.5-cm^2^ surface electrodes placed approximately 2 cm apart, with a ground electrode on the lateral condyle. 
-For this preliminary analysis, we considered only the activity of tibialis anterior of the right leg during forward perturbations.
+The activity of tibialis anterior (TA) and medial gastrocnemius (MG) were recorded.
+For this analysis, we considered only the activity of tibialis anterior of the right leg.
 
+***
 ```{r perturbationFigure, out.width = "25%", fig.align = "left", fig.cap = captionCartoon}
@@
 ```
+***
 
@@
 
+taR = taR %>% mutate(Group = factor(`Study Group`, levels = c("Control", "MS"))) %>% select(-`Study Group`) %>% 
+  mutate(`Perturbation Direction` = factor(`Perturbation Direction`, levels = c("Forward", "Backward")))
+
+# calculate max activity for normalization purposes. this has already been filtered, etc. so it is probably ok just to normalize to maximum values.
+taRNorm = taR %>% group_by(Participant) %>% summarize(Vm = max(V, na.rm = TRUE)) %>% ungroup()
+taR = taR %>% left_join(taRNorm, by = "Participant") %>% mutate(Vn = V/Vm)
+
 ```
 
+# Results
+
+## Demographic and clinical characteristics
+Demographic and clinical characteristics of the study population are summarized in `r tableNumbers("demographics", display = "cite")`.
+
+***
 ```{r demographics}
@@
 ```
+***
 
-```{r rawEmg}
+## Changes in tibialis anterior activity with disease state and perturbation direction
+Recorded EMG traces are presented in `r figureNumbers("rawEmg", display = "cite")`.
+
+***
+```{r rawEMG, out.width = "100%", fig.align = "left", fig.cap = captionRawEmg}
 library(ggplot2)
-p = ggplot(taR, aes(x = Time, y = V, group = Code)) + geom_line() + facet_grid(`Study Group` ~ `Perturbation Direction`)
+library(cowplot)
+p = ggplot(taR, aes(x = Time, y = Vn, group = Code)) + geom_line() + facet_grid(Group ~ `Perturbation Direction`) + theme_minimal_grid()
 p
 ```
+***
 
-## Demographic and clinical characteristics
-Demographic and clinical characteristics of the study population are summarized in `r tableNumbers("demographics", display = "cite")`.
+# Discussion
+
+TK
 

### 2020-02-04 R/msBalance.Rmd
added notes about normalization. probably better to average across replicates prior to calculating max values, will be less sensitive to high frequency data.
additions to plot normalization

@@
 library(lubridate)
+library(janitor)
+library(kableExtra)
+library(ggplot2)
+library(cowplot)
 `%contains%` = function(s, pattern) grepl(pattern, s)
 `%omits%` = function(s, pattern) !grepl(pattern, s)
+`%between%` = function(s, window) s>=min(window)&s<max(window)
 
@@
 tableNumbers <- captioner(prefix = "Table")
+captionEmgMags = tableNumbers(name = "emgMags", caption = "Maximum EMG levels.")
 captionDemographics = tableNumbers(name = "demographics", caption = "Clinical and demographic features of the study population.")
+
 figureNumbers <- captioner(prefix = "Figure")
 captionCartoon = figureNumbers(name = "cartoon", caption = "Perturbation directions used in testing protocol.")
-captionRawEmg = figureNumbers(name = "rawEmg", caption = "EMG traces stratified by perturbation direction and study group.")
+captionRawEmg = figureNumbers(name = "rawEmg", caption = "EMG traces stratified by perturbation direction and study group, prior to normalization. Colors represent individual participants.")
 ```
@@
 Each individual EMG record was registered to the onset of treadmill motion and resampled to a common sample rate (1 kHz), and processed (high-pass, 35 Hz, de- mean, rectify, low-pass, 40 Hz) as in previous studies.[@lang2019; @mckay2016]
+Recorded EMG records were visualized stratified by disease state and perturbation direction (`r figureNumbers("rawEmg", display = "cite")`).
+By visual inspection, there were some gross magnitude differences in recorded EMG levels between the MS and control groups, with higher EMG levels overall in the MS group.
+EMG levels were therefore normalized to the maximum values obtained in a 1 second window after perturbation onset after averaging across like trials (`r tableNumbers("emgMags", display = "cite")`).
 
@@
 # I cannot figure out how to get dplyr::left_join to join by a variable with backticks, so rename temporarily
-taR = left_join(taR, trialData, by = "Code") %>% 
+taR %<>% left_join(trialData, by = "Code") %>% 
   select(`Study Group`,Participant, Date, Trial, `Code`, `Trial Number`, `Perturbation Direction`, Muscle, Side, everything()) %>% 
@@
 # note that a large amount of the data (40%) is follow-up, with different visit amounts. for now, let's just consider the initial visit.
-taR = taR %>% filter(Visit == "Initial")
+taR %<>% filter(Visit == "Initial")
 
@@
 includedParticipants = intersect(taR$Participant, participants$Participant)
-taR = taR %>% filter(Participant %in% includedParticipants)
-participants = participants %>% filter(Participant %in% includedParticipants)
 
-taR = taR %>% mutate(Group = factor(`Study Group`, levels = c("Control", "MS"))) %>% select(-`Study Group`) %>% 
+# consider only a subset of participants if debugging
+SP = FALSE
+if (SP){
+  includedParticipants = sample(includedParticipants, 60)
+}
+
+taR %<>% filter(Participant %in% includedParticipants)
+participants %<>% filter(Participant %in% includedParticipants)
+
+# here are the time records
+taR %<>% mutate(Group = factor(`Study Group`, levels = c("Control", "MS"))) %>% select(-`Study Group`) %>% 
   mutate(`Perturbation Direction` = factor(`Perturbation Direction`, levels = c("Forward", "Backward")))
 
-# calculate max activity for normalization purposes. this has already been filtered, etc. so it is probably ok just to normalize to maximum values.
-taRNorm = taR %>% group_by(Participant) %>% summarize(Vm = max(V, na.rm = TRUE)) %>% ungroup()
-taR = taR %>% left_join(taRNorm, by = "Participant") %>% mutate(Vn = V/Vm)
+# here are the averages across like trials
+taRAv = taR %>% group_by(Group, Participant, Muscle, Side, `Perturbation Direction`, Time) %>% summarize(V = mean(V, na.rm = TRUE)) %>% ungroup()
 
+# to calculate normalization coefficients, average across like trials for each patient, limit to a particular normalization window, and calculate maximum values
+taRNorm = taRAv %>% filter(Time %between% c(0.0, 1.0)) %>% group_by(Group, Participant, `Perturbation Direction`) %>% summarize(Vm = max(V, na.rm = TRUE)) %>% ungroup()
 ```
+
+***
+```{r rawEMG, out.width = "100%", fig.align = "left", fig.cap = captionRawEmg}
+p = ggplot(taR, aes(x = Time, y = V)) + geom_line(aes(color = Participant), show.legend = FALSE, alpha = 0.5) + facet_grid(Group ~ `Perturbation Direction`) + theme_minimal_grid() +
+  geom_jitter(mapping = aes(x = 0, y = Vm, color = Participant), data = taRNorm %>% left_join(participants %>% select(Participant, Group)), show.legend = FALSE, height = 0, width = 0.01)
+p
+```
+
+```{r emgMags}
+emgMags = 
+  taRNorm %>% group_by(`Perturbation Direction`, Group) %>% summarize(Mean = mean(Vm) %>% round() ) %>% left_join(
+    taRNorm %>% group_by(`Perturbation Direction`, Group) %>% summarize(SD = sd(Vm) %>% round() ) ) %>% left_join(
+      taRNorm %>% group_by(`Perturbation Direction`, Group) %>% summarize(Max = max(Vm) %>% round() ) )
+emgMags %>% kable(caption = captionEmgMags) %>% kable_styling(full_width = F, position = "left")
+
+# 
+# 
+# ggplot(taRAv, aes(Time, Vn, color = Participant)) + geom_point(show.legend = FALSE) + facet_wrap(vars(`Perturbation Direction`))
+# 
+# p = ggplot(taRAv, aes(Time, Vn)) + geom_point()
+# 
+# p = ggplot(taRAv %>% group_by(Participant, `Perturbation Direction`), aes(x = Time, y = Vn, color = Participant)) +
+#   facet_grid(Group ~ `Perturbation Direction`) +
+#   theme_minimal_grid() +
+#   coord_cartesian(xlim = c(0,1)) + 
+#   geom_line(aes(group = Participant, color = Participant), taRav, show.legend = FALSE) +
+# p
+```
+
+***
 
@@
 ```{r demographics}
-library(janitor)
-library(kableExtra)
 participants %>% tabyl(Sex, Group) %>% mutate(Total = MS + Control) %>% 
@@
 ## Changes in tibialis anterior activity with disease state and perturbation direction
-Recorded EMG traces are presented in `r figureNumbers("rawEmg", display = "cite")`.
-
-***
-```{r rawEMG, out.width = "100%", fig.align = "left", fig.cap = captionRawEmg}
-library(ggplot2)
-library(cowplot)
-p = ggplot(taR, aes(x = Time, y = Vn, group = Code)) + geom_line() + facet_grid(Group ~ `Perturbation Direction`) + theme_minimal_grid()
-p
-```
-***
 

### 2020-02-04 R/msBalance.Rmd
added labels to normalization figure

@@
 ```{r rawEMG, out.width = "100%", fig.align = "left", fig.cap = captionRawEmg}
+# p = ggplot(taR, aes(x = Time, y = V)) + geom_line(aes(color = Participant), show.legend = FALSE, alpha = 0.5) + facet_grid(Group ~ `Perturbation Direction`) + theme_minimal_grid() +
+#   geom_jitter(mapping = aes(x = 0, y = Vm, color = Participant), data = taRNorm %>% left_join(participants %>% select(Participant, Group)), show.legend = FALSE, height = 0, width = 0.01)
+# p
 p = ggplot(taR, aes(x = Time, y = V)) + geom_line(aes(color = Participant), show.legend = FALSE, alpha = 0.5) + facet_grid(Group ~ `Perturbation Direction`) + theme_minimal_grid() +
-  geom_jitter(mapping = aes(x = 0, y = Vm, color = Participant), data = taRNorm %>% left_join(participants %>% select(Participant, Group)), show.legend = FALSE, height = 0, width = 0.01)
+  geom_label(mapping = aes(x = 0, y = Vm, color = Participant, label = Participant), data = taRNorm, show.legend = FALSE)
 p

### 2020-02-04 R/msBalance.Rmd
updated plots of EMG magnitudes to show only the top 3 codes of each cell

--- 2020-02-04 R/msBalance.Rmd
@@
 Recorded EMG records were visualized stratified by disease state and perturbation direction (`r figureNumbers("rawEmg", display = "cite")`).
-By visual inspection, there were some gross magnitude differences in recorded EMG levels between the MS and control groups, with higher EMG levels overall in the MS group.
-EMG levels were therefore normalized to the maximum values obtained in a 1 second window after perturbation onset after averaging across like trials (`r tableNumbers("emgMags", display = "cite")`).
 
@@
 ```{r rawEMG, out.width = "100%", fig.align = "left", fig.cap = captionRawEmg}
-p = ggplot(taR, aes(x = Time, y = V)) + geom_line(aes(color = Participant), show.legend = FALSE, alpha = 0.5) + facet_grid(Group ~ `Perturbation Direction`) + theme_minimal_grid() +
-  geom_jitter(mapping = aes(x = 0, y = Vm, color = Participant), data = taRNorm %>% left_join(participants %>% select(Participant, Group)), show.legend = FALSE, height = 0, width = 0.01)
+# p = ggplot(taR, aes(x = Time, y = V)) + geom_line(aes(color = Participant), show.legend = FALSE, alpha = 0.5) + facet_grid(Group ~ `Perturbation Direction`) + theme_minimal_grid() +
+#   geom_jitter(mapping = aes(x = 0, y = Vm, color = Participant), data = taRNorm %>% left_join(participants %>% select(Participant, Group)), show.legend = FALSE, height = 0, width = 0.01)
+# p
+# p = ggplot(taR, aes(x = Time, y = V)) + geom_line(aes(color = Participant), show.legend = FALSE, alpha = 0.5) + facet_grid(Group ~ `Perturbation Direction`) + theme_minimal_grid() +
+#   geom_label(mapping = aes(x = 0, y = Vm, color = Participant, label = Participant), data = taRNorm, show.legend = FALSE)
+# p
+p = ggplot(taR, aes(x = Time, y = V)) +
+  geom_line(color = "Gray90", show.legend = FALSE, alpha = 0.5) +
+  geom_line(aes(color = Participant), taRAv, show.legend = FALSE, alpha = 0.5) +
+  facet_grid(Group ~ `Perturbation Direction`) +
+  theme_minimal_grid() +
+  geom_text(mapping = aes(x = 0, y = Vm, color = Participant, label = Participant), data = taRNorm %>% group_by(Group, `Perturbation Direction`) %>% top_n(3), show.legend = FALSE)
 p
 ```
+
+By visual inspection, there were some gross magnitude differences in recorded EMG levels between the MS and control groups, with higher EMG levels overall in the MS group.
+EMG levels in most control participants (with the exception of MSC15 and MSC18) were generally roughly an order of magnitude lower than levels in MS participants.
+This number is difficult to interpret, as it can reflect a large number of biological and methodological factors, so that increased absolute level among MS could reflect increased muscle activation or completely unrelated factors.
+EMG levels were therefore normalized to the maximum values obtained in a 1 second window after perturbation onset after averaging across like trials (`r tableNumbers("emgMags", display = "cite")`).
 

### 2020-02-04 R/msBalance.Rmd
changed name of norm coef tibble from taRNorm to taRMax

--- 2020-02-04 R/msBalance.Rmd
@@
 captionCartoon = figureNumbers(name = "cartoon", caption = "Perturbation directions used in testing protocol.")
-captionRawEmg = figureNumbers(name = "rawEmg", caption = "EMG traces stratified by perturbation direction and study group, prior to normalization. Colors represent individual participants.")
+captionRawEmg = figureNumbers(name = "rawEmg", caption = "EMG traces stratified by perturbation direction and study group, prior to normalization.
+Gray traces represent individual trials.
+Colors represent averages of individual participants.")
+captionNormedEmg = figureNumbers(name = "normedEmg", caption = "EMG traces stratified by perturbation direction and study group, after normalization.
+Colors represent averages for individual participants.")
 ```
@@
 # to calculate normalization coefficients, average across like trials for each patient, limit to a particular normalization window, and calculate maximum values
-taRNorm = taRAv %>% filter(Time %between% c(0.0, 1.0)) %>% group_by(Group, Participant, `Perturbation Direction`) %>% summarize(Vm = max(V, na.rm = TRUE)) %>% ungroup()
+taRMax = taRAv %>% filter(Time %between% c(0.0, 1.0)) %>% group_by(Group, Participant, `Perturbation Direction`) %>% summarize(Vm = max(V, na.rm = TRUE)) %>% ungroup()
 ```
@@
 ***
-```{r rawEMG, out.width = "100%", fig.align = "left", fig.cap = captionRawEmg}
-p = ggplot(taR, aes(x = Time, y = V)) + geom_line(aes(color = Participant), show.legend = FALSE, alpha = 0.5) + facet_grid(Group ~ `Perturbation Direction`) + theme_minimal_grid() +
-  geom_jitter(mapping = aes(x = 0, y = Vm, color = Participant), data = taRNorm %>% left_join(participants %>% select(Participant, Group)), show.legend = FALSE, height = 0, width = 0.01)
+```{r rawEMG, out.width = "75%", fig.align = "left", fig.cap = captionRawEmg}
+p = ggplot(taR, aes(x = Time, y = V)) +
+  geom_line(color = "Gray90", show.legend = FALSE, alpha = 0.5) +
+  geom_line(aes(color = Participant), taRAv, show.legend = FALSE, alpha = 0.5) +
+  facet_grid(Group ~ `Perturbation Direction`) +
+  theme_minimal_grid() +
+  geom_text(mapping = aes(x = 0, y = Vm, color = Participant, label = Participant), data = taRMax %>% group_by(Group, `Perturbation Direction`) %>% top_n(3), show.legend = FALSE)
 p
@@
 
+By visual inspection, there were some gross magnitude differences in recorded EMG levels between the MS and control groups, with higher EMG levels overall in the MS group.
+EMG levels in most control participants (with the exception of MSC15 and MSC18) were generally roughly an order of magnitude lower than levels in MS participants.
+This number is difficult to interpret, as it can reflect a large number of biological and methodological factors, so that increased absolute level among MS could reflect increased muscle activation or completely unrelated factors.
+EMG levels were therefore normalized to the maximum values obtained in a 1 second window after perturbation onset after averaging across like trials (`r tableNumbers("emgMags", display = "cite")`).
+
 ```{r emgMags}
 emgMags = 
-  taRNorm %>% group_by(`Perturbation Direction`, Group) %>% summarize(Mean = mean(Vm) %>% round() ) %>% left_join(
-    taRNorm %>% group_by(`Perturbation Direction`, Group) %>% summarize(SD = sd(Vm) %>% round() ) ) %>% left_join(
-      taRNorm %>% group_by(`Perturbation Direction`, Group) %>% summarize(Max = max(Vm) %>% round() ) )
+  taRMax %>% group_by(`Perturbation Direction`, Group) %>% summarize(Mean = mean(Vm) %>% round() ) %>% left_join(
+    taRMax %>% group_by(`Perturbation Direction`, Group) %>% summarize(SD = sd(Vm) %>% round() ) ) %>% left_join(
+      taRMax %>% group_by(`Perturbation Direction`, Group) %>% summarize(Max = max(Vm) %>% round() ) )
 emgMags %>% kable(caption = captionEmgMags) %>% kable_styling(full_width = F, position = "left")
-
-# 
-# 
-# ggplot(taRAv, aes(Time, Vn, color = Participant)) + geom_point(show.legend = FALSE) + facet_wrap(vars(`Perturbation Direction`))
-# 
-# p = ggplot(taRAv, aes(Time, Vn)) + geom_point()
-# 
-# p = ggplot(taRAv %>% group_by(Participant, `Perturbation Direction`), aes(x = Time, y = Vn, color = Participant)) +
-#   facet_grid(Group ~ `Perturbation Direction`) +
-#   theme_minimal_grid() +
-#   coord_cartesian(xlim = c(0,1)) + 
-#   geom_line(aes(group = Participant, color = Participant), taRav, show.legend = FALSE) +
-# p
 ```
 
-***
+```{r normedEMG, out.width = "75%", fig.align = "left", fig.cap = captionNormedEmg}
+p = ggplot(taR, aes(x = Time, y = V)) +
+  geom_line(color = "Gray90", show.legend = FALSE, alpha = 0.5) +
+  geom_line(aes(color = Participant), taRAv, show.legend = FALSE, alpha = 0.5) +
+  facet_grid(Group ~ `Perturbation Direction`) +
+  theme_minimal_grid() +
+  geom_text(mapping = aes(x = 0, y = Vm, color = Participant, label = Participant), data = taRMax %>% group_by(Group, `Perturbation Direction`) %>% top_n(3), show.legend = FALSE)
+p
+```
 

### 2020-02-04 R/msBalance.Rmd
added code to plot normalized EMG

@@
 taRMax = taRAv %>% filter(Time %between% c(0.0, 1.0)) %>% group_by(Group, Participant, `Perturbation Direction`) %>% summarize(Vm = max(V, na.rm = TRUE)) %>% ungroup()
+
+# take the maximum observed in either the forward or backward directions.
+taRMaxMax = taRMax %>% group_by(Participant) %>% summarize(Vm = max(Vm))
+
+# append normed EMG to existing EMG
+taR %<>% left_join(taRMaxMax, by = "Participant") %>% mutate(Vn = V / Vm)
+taRAv %<>% left_join(taRMaxMax, by = "Participant") %>% mutate(Vn = V / Vm)
 ```
@@
 ```{r normedEMG, out.width = "75%", fig.align = "left", fig.cap = captionNormedEmg}
-p = ggplot(taR, aes(x = Time, y = V)) +
+p = ggplot(taR, aes(x = Time, y = Vn)) +
   geom_line(color = "Gray90", show.legend = FALSE, alpha = 0.5) +
@@
   theme_minimal_grid() +
-  geom_text(mapping = aes(x = 0, y = Vm, color = Participant, label = Participant), data = taRMax %>% group_by(Group, `Perturbation Direction`) %>% top_n(3), show.legend = FALSE)
+  coord_cartesian(ylim = c(0,2))
 p

### 2020-02-14 R/msBalance.Rmd
added sample markdown
Update msBalance.Rmd

@@
 The main objective of this study was to examine the impact of MS on the pattern of magnitude and timing of muscle activation during feet-in-place balance responses.
+
+Previous work suggests that people with MS may exibit delays in the release of reactive postural control movements (Cameron 2008, Huisinga 2014, Peterson 2015, etc.).
+However, little work has established the degree to which other aspects of the postural response are or are not impacted in people with MS.
+For example, in seminal but small-sample (6 MS participants) early work, Cameron et al.
+2008 showed that people with MS exhibited larger postural responses (measured via torque underfoot) compared to neurotypical adults.
+They suggested that this may have been an effort to compensate for the delayed response onset.
+While these data provide a glimpse into the impacts of MS on automatic postural responses, additional work is necessary to understand how MS impacts the timing, amplitude, and length of automatic postural responses.
+Further, characterizing both agonists and antagonist activity can interrogate potential coactivation of postural repsonse muscles during a loss of balance.
+indeed, We (McKay) have observed increased coactivation in other neurogical populations in this circumstance.
+
+[Backward losses of balance can be particularly difficult and injurious in neurological populations (Horak & Dimitrova 2004, Carpenter).
+Interestingly, recent work has suggested that these two movements may rely on slightly different neural structures (Nonnekes et al.
+2013; Neuroscience).
+Although considerably more work is necessary to establish this hypothesis, it does underscore the importance of assessing both forward and backward reactive postural control, especially in neurological populations.]
+
+Therefore, the goal of this project is to characterize muscular responses to an external perturbation in a large sample of MS participants.
+Based on previous results, we hypothesize that agonist EMG activation will be delayed and larger in peopel with MS compared to neurotypical peers.
+[Further, we expect increased activity of antagonists.
+We expect these deficits to 1) be related to retrospective and pro(ish)-spective fall history.
+Finally, we expect these deficits to become more pronounced over the 18 month observation period.]
+
+## Subheading
+
+This is some markdown text.
+1. Bulleted lists are easy.
+1. For real.
 
@@
 Complete details of the assessment protocol are avaialble in an earlier report.[@huisinga2014] 
-During each asssessment, subjects stood with arms folded across the chest, eyes open, with their feet at a fixed heel-to-heel distance of 10 cm on a set of 2 custom hydraulic force platforms that translated either forward or backward (`r figureNumbers("cartoon", display = "cite")`).
+Participant demographics (age, year of diagnosis, medications) as well as  clinical and balance assessments (EDSS, Berg Balance scale, Timed Up and Go, 25ft walk test) were completed.
+Retrospective falls over the 6 months prior to study entry was also queried.
+Participants were then completed a reactive balance assessment.
+Subjects stood with arms folded across the chest, eyes open, with their feet at a fixed heel-to-heel distance of 10 cm on a set of 2 custom hydraulic force platforms that translated either forward or backward (`r figureNumbers("cartoon", display = "cite")`).
 In this study, "perturbation direction" refers to the direction of postural sway caused by the surface translation.
@@
 
-## Electromyography
+[Participants completed 4 visits to the laboratory, each separated by approximately 6 months.
+Reactive stepping assessments were completed at each timepoint (Notably, there was very high attrition, that will be discussed in the results.))]
+
+## Data collection
 Surface electromyography was recorded with pairs 2.5-cm^2^ surface electrodes placed approximately 2 cm apart, with a ground electrode on the lateral condyle. 
@@
 For this analysis, we considered only the activity of tibialis anterior of the right leg.
+
+A reflective marker (tracked by Motion Analysis Corp cameras) was also placed on the treadmill belt to capture support surface movement and for calculation of perturbation onset.
 
